# Standard normal probability distribution {#app:criticalZvalues}

<!-- kritiekeZwaarden_hulpfiguur.R, no caption, no labels -->
```{r kritiekeZwaarden-hulpfiguur, echo=FALSE, fig.height=2}
op <- par(mar=c(3,1,1,1)+0.1)
xx <- seq(1,4,by=.01)
curve( dnorm(x), from=-4, to=4, xaxt="n", yaxt="n", xlab="", ylab="", bty="n", lwd=3 )
yy <- dnorm(xx,0,1)
polygon( x=c(1,xx), y=c(0,yy), col="lavender", border="lavender" )
# 20201128 added other polygon corresponding to B, three commands
xx <- seq(-1,1,by=.01) 
yy <- dnorm(xx,0,1)
polygon( x=c(-1,xx,1), y=c(0,yy,0), col="wheat1", border="wheat1" )
#
curve( dnorm(x), from=-4, to=4, xaxt="n", yaxt="n", xlab="", ylab="", bty="n", lwd=3, add=T )
text( 1.5, dnorm(1.5)*.4, "p", adj=.5, cex=1 )
# 20201128 added other polygon corresponding to B, two more commands
text( -0.5, dnorm(0.5)*.4, "B/2", adj=.5, cex=1 )
text( +0.5, dnorm(0.5)*.4, "B/2", adj=.5, cex=1 )
axis( side=1, at=c(0,1), labels=c("0","Z*"), cex=1, tick=F, line=-1 ) # cex is ignored
# abline( h=0, lty=1, col="grey70" ) # should not extend outside tails of distribution
segments( x0=-4, x1=4, y0=0, col="grey70" ) # as X axis
segments( x0=0, x1=0, y0=0, y1=dnorm(0), col="grey70" )
segments( x0=1, x1=1, y0=0, y1=yy[1], col="grey70" )
curve( dnorm(x), from=-4, to=4, xaxt="n", yaxt="n", xlab="", ylab="", bty="n", lwd=3, add=T )
par(op)
```

The critical value $Z^*$ given below has a probability of $p$ under $H_0$, i.e., 
$P(Z > Z^*|H_0)=p$ (the blue area), 
and it has a probability of $B$ to have a value in the interval $(-Z^*, +Z^*)$ (the yellow area). 
The $Z$ distribution is symmetrical around $Z=0$, hence $P(Z < -Z^*) = P(Z > Z^*)$.

---

The first table reports the critical boundary values $Z^*$ for some frequently used probabilities of $p$ and frequently used confidence intervals of $B$:

```{r A2.1, echo=FALSE}
pvals <- c(.2,.1,.05,.025,.01,.005,.0025,.001)
appendixZ1 <- matrix("",nrow=1,ncol=1+length(pvals)) # matrix of char strings
for (i in 1:1) 
  for (j in 1:length(pvals))
    appendixZ1[i,j+1] <- format( qnorm( p=pvals[j], lower.tail=FALSE),
                               digits=4, nsmall=3 )
appendixZ1 <- rbind(
  c( "p", as.character(pvals) ),
  c( "B", paste(as.character(100-200*pvals),"%",sep="" ) ),
  appendixZ1
)
appendixZ1[3,1] <- "Z*"
# suppress first row, and use as column names
knitr::kable(
  appendixZ1[-1,], booktabs=TRUE, col.names=appendixZ1[1,], align="c" )
```

---

The second table reports the probabilities $p$ and confidence intervals $B$ for some frequently used critical values of $Z^*$:

```{r A2.2, echo=FALSE}
Zvals <- seq(from=0.5, to=3.5, by=0.5)
pvals <- pnorm(Zvals, lower.tail=FALSE )
appendixZ2 <- matrix("",nrow=1,ncol=length(pvals)+1)
# for (i in 1:1) 
#  for (j in 1:length(pvals))
#    appendixZ2[i,j+1] <- format( Zvals,
#                               digits=3, nsmall=1 )
appendixZ2[1,2:ncol(appendixZ2)] <- Zvals
appendixZ2 <- rbind(
  c( "p", format(pvals, digits=1, nsmall=4) ),
  c( "B", paste(as.character(round(100-200*pvals,digits=2)),"%",sep="" ) ),
  appendixZ2
)
appendixZ2[3,1] <- "Z*"
# suppress first row, and use as column names
knitr::kable(
  appendixZ2[-1,], booktabs=TRUE, col.names=appendixZ2[1,], align="c" )
```
